Optimal synaptic strategies for different timescales of memory

### 2018-03-16 Paper/SynapsePaper.tex
new pnas doc class

%% PNAStwoS.tex
%% For two column PNAS articles
%% Version: Apr 15, 2008

%% BASIC CLASS FILE
\documentclass{pnastwo}

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage{graphicx}
\usepackage{ifpdf}
%\usepackage[hypcap]{caption}%links to figures
\usepackage{graphicx}
%\usepackage{graphics} for finding documentation only
%
\usepackage{epstopdf}
\epstopdfsetup{update,suffix=-generated}
%
%\newcommand{\inkscapedir}{C:/Program Files/Inkscape/}
%\IfFileExists{"C:/Program Files (x86)/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Program Files (x86)/Inkscape/}}{}
\IfFileExists{"C:/Program Files/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Program Files/Inkscape/}}{}
\IfFileExists{"/Applications/Inkscape.app/Contents/Resources/bin/inkscape"}{\newcommand{\inkscapedir}{/Applications/Inkscape.app/Contents/Resources/bin/}}{}
%
\ifpdf
 \epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{%
 "\inkscapedir inkscape" -f #1 -C -A \OutputFile
 }
\else
 \DeclareGraphicsRule{.png}{eps}{.bb}{}
\fi
%
\usepackage{adjustbox}
\newcommand{\aligntop}[1]{\adjustbox{valign=t}{#1}}
\newcommand{\alignmid}[1]{\adjustbox{valign=c}{#1}}
%
%\usepackage{paralist}
%\newenvironment{myenuma}{\begin{inparaenum}[(a)]}{\end{inparaenum}}
%\newenvironment{myenumi}{\begin{inparaenum}[i)]}{\end{inparaenum}}
%\newenvironment{myenumA}{\begin{inparaenum}[\bfseries A.]}{\end{inparaenum}}
%\newenvironment{myenumI}{\begin{inparaenum}[\bfseries I.]}{\end{inparaenum}}
%\newenvironment{myenumr}{\begin{inparaenum}[1.]}{\end{inparaenum}}
%%for nested lists, \ref only produces innermost item:  (hacky)
%\makeatletter % for internal macros with @
%\renewcommand\p@enumii{}
%\makeatother
%
\usepackage[inline]{enumitem}
\newlist{myenuma}{enumerate*}{4}
\setlist[myenuma]{label=(\alph*),ref=\alph*}
%\makeatletter
%\DeclareRobustCommand{\ilabel}[1]{%
%   \edef\@currentlabel{\themyenumai}%
%   \label{#1}}
%\makeatother
\graphicspath{{"Figs/"}}
%\usepackage{pnastwoF}
%\usepackage{amssymb,amsfonts}
%\usepackage[centertags]{amsmath}
%\usepackage{hyperref}

%% OPTIONAL MACRO DEFINITIONS
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
% indices
%
% brackets etc.
\newcommand{\prn}[1]{\left ( #1 \right )}
\newcommand{\brc}[1]{\left\{ #1 \right\}}
\newcommand{\brk}[1]{\left [ #1 \right ]}
\newcommand{\abs}[1]{\left\lvert #1 \right\rvert}
\newcommand{\nrm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\avc}[2]{\left\langle #1 \,\middle | \, #2 \right\rangle}
%\newenvironment{cases}{\left\{\begin{aligned}}{\end{aligned}\right.}
%
% Sets
\newcommand{\set}[2]{\left\{ #1 \,\middle |\, #2 \right\}}
\newcommand{\cond}[2]{\!\left( #1 \,\middle | \, #2 \right)}
\newcommand{\condb}[2]{\!\left[ #1 \,\middle | \, #2 \right]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][\rule{0mm}{0mm}]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][\rule{0mm}{0mm}]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][\rule{0mm}{0mm}]{\left (\frac{\partial #2}{\partial {#3}}\right )_{\!\!#1}}
\newcommand{\intd}[2][\rule{0mm}{0mm}]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
% Referencing sections, figures, etc
\newcommand{\sref}[1]{\S\ref{#1}}
\newcommand{\fref}[1]{fig.\ref{#1}}
\newcommand{\Fref}[1]{Fig.~\ref{#1}}
\newcommand{\tref}[1]{tab.\ref{#1}}
\newcommand{\Tref}[1]{Tab.\ref{#1}}
%
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\CL}{\mathcal{L}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

% ----------------------------------------------------------------
\input{sl_theorems_preamble.tex}
% ----------------------------------------------------------------
%\renewcommand\Fref[1]{}
\providecommand\hyperref[2][1]{#2}
\providecommand\texorpdfstring[2]{#1}

%%%%%%%%%%%%
%% For PNAS Only:
%\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\renewcommand{\theurl}{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2016}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%\setcounter{page}{2687} %Set page number here if desired
%%%%%%%%%%%%

\begin{document}

\title{Optimal synaptic strategies for different timescales of memory}

\author{Subhaneil Lahiri\affil{1}{Department of Applied Physics, Stanford University, Stanford CA}
\and
Surya Ganguli\affil{1}{}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

\maketitle

\begin{article}
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

@@

\keywords{synaptic plasticity | learning | memory | neuroscience}

\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation;
LTD, long term depression}

@@

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \Fref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

We use a well established formalism for the study of learning and memory with complex synapses (see \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}).
In this approach, electrical patterns of activity corresponding to candidate potentiating and depressing plasticity events occur randomly and independently at all synapses at a Poisson rate $r$.  These events reflect possible synaptic changes due to either spontaneous network activity, or the storage of new memories.
We let  $f\pot$ and  $f\dep$ denote the fraction of these events that are candidate potentiating or depressing events respectively.
Furthermore, we assume our synaptic model has $M$ internal molecular functional states, and that a candidate potentiating (depotentiating) event induces a stochastic transition in the internal state described by an $M \times M$ discrete time Markov transition matrix  $\M\pot$ ($\M\dep$).
In this framework, the states of different synapses will be independent, and the entire synaptic population can be fully described by the probability distribution across these states, which we will indicate with the row-vector $\pr(t)$.
Thus the $i$'th component of $\pr(t)$ denotes the fraction of the synaptic population in state $i$.
Furthermore, each state $i$ has its own synaptic weight, $\w_i$, which we take, in the worst case scenario, to be restricted to two values.
After shifting and scaling these two values,  we can assume they are $\pm1$, without loss of generality.

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \Fref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  \label{fig:models}}
\end{figure}

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \eqref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \eqref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[sec:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

\begin{acknowledgments}
We thank Sloan, Genenetech, Burroughs-Wellcome, and Swartz foundations for support.
We thank Larry Abbott, Marcus Benna, Stefano Fusi, Jascha Sohl-Dickstein and David Sussillo for useful discussions.
\end{acknowledgments}

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage[hypcap]{caption}%links to figures
\usepackage{graphicx}
%\usepackage{graphics} for finding documentation only
\pdfsuppresswarningpagegroup=1
\usepackage{epstopdf}
\epstopdfsetup{update,suffix=-generated}
%
\IfFileExists{"C:/Progra\string~1/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Progra\string~1/Inkscape/}}{}
\IfFileExists{"/Applications/Inkscape.app/Contents/Resources/bin/inkscape"}{\newcommand{\inkscapedir}{/Applications/Inkscape.app/Contents/Resources/bin/}}{}
\providecommand{\inkscapedir}{}
%
\epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{"\inkscapedir inkscape" -f #1 -C -A \OutputFile}
%
\usepackage{adjustbox}
\newcommand{\aligntop}[1]{\adjustbox{valign=t}{#1}}
%
%\usepackage{paralist}
%\newenvironment{myenuma}{\begin{inparaenum}[(a)]}{\end{inparaenum}}
%\newenvironment{myenumA}{\begin{inparaenum}[\bfseries A.]}{\end{inparaenum}}
%
%\usepackage[inline]{enumitem}
\newlist{myenuma}{enumerate*}{4}
\setlist[myenuma]{label=(\alph*),ref=\alph*}
\newcommand{\itemp}[1]{\item\aligntop{#1}}

\graphicspath{{"Figs/"}}
\usepackage{amssymb,amsfonts}
%\usepackage[centertags]{amsmath}
%\usepackage{hyperref}
\usepackage[capitalise,nameinlink]{cleveref}
\creflabelformat{equation}{[#2#1#3]}
\crefrangelabelformat{equation}{[#3#1#4--#5#2#6]}
%\crefformat{equation}{#2Eq.~[#1]#3}
%\crefrangeformat{equation}{Eqs.~[#3#1#4--#5#2#6]}
%\crefmultiformat{equation}{Eqs.~[#2#1#3}{,2#1#3]}{,#2#1#3}{,#2#1#3]}
%\crefrangemultiformat{equation}{Eqs.~[#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}{,#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}

%% OPTIONAL MACRO DEFINITIONS
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
\usepackage{delim}
% brackets etc.
\delimdef\prn#1{\dleft( #1 \dright)}
\delimdef\brc#1{\dleft\{ #1 \dright\}}
\delimdef\brk#1{\dleft[ #1 \dright]}
\delimdef\abs#1{\dleft\lvert #1 \dright\rvert}
\delimdef\nrm#1{\dleft\lVert #1 \dright\rVert}
\delimdef\av#1{\dleft\langle #1 \dright\rangle}
\delimdef\avc#1#2{\dleft\langle #1 \dmiddle\vert #2 \dright\rangle}
%\newenvironment{cases}{\dleft\{\begin{aligned}}{\end{aligned}\dright.}
%
% Sets
\delimdef\set#1#2{\dleft\{ #1 \dmiddle\vert #2 \dright\}}
\delimdef\cond#1#2{\!\dleft( #1 \dmiddle\vert #2 \dright)}
\delimdef\condb#1#2{\!\dleft[ #1 \dmiddle\vert #2 \dright]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][]{\left(\frac{\partial #2}{\partial {#3}}\right)_{\!\!#1}}
\newcommand{\intd}[2][]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\CL}{\mathcal{L}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

% ----------------------------------------------------------------
\input{sl_theorems_preamble.tex}
% ----------------------------------------------------------------
\providecommand\hyperref[2][1]{#2}
\providecommand\texorpdfstring[2]{#1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Optimal synaptic strategies for different timescales of memory}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1,2]{Subhaneil Lahiri}
\author[b,1]{Surya Ganguli}

\affil[a]{Department of Applied Physics, Stanford University, Stanford CA}

% Please give the surname of the lead author for the running footer
\leadauthor{Lahiri}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sulahiri@stanford.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{synaptic plasticity | learning | memory | neuroscience}

%\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation;LTD, long term depression}

%\begin{article}
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

@@

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{article}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \cref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

We use a well established formalism for the study of learning and memory with complex synapses (see \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}).
In this approach, electrical patterns of activity corresponding to candidate potentiating and depressing plasticity events occur randomly and independently at all synapses at a Poisson rate $r$.  These events reflect possible synaptic changes due to either spontaneous network activity, or the storage of new memories.
We let  $f\pot$ and  $f\dep$ denote the fraction of these events that are candidate potentiating or depressing events respectively.
Furthermore, we assume our synaptic model has $M$ internal molecular functional states, and that a candidate potentiating (depotentiating) event induces a stochastic transition in the internal state described by an $M \times M$ discrete time Markov transition matrix  $\M\pot$ ($\M\dep$).
In this framework, the states of different synapses will be independent, and the entire synaptic population can be fully described by the probability distribution across these states, which we will indicate with the row-vector $\prob(t)$.
Thus the $i$'th component of $\prob(t)$ denotes the fraction of the synaptic population in state $i$.
Furthermore, each state $i$ has its own synaptic weight, $\w_i$, which we take, in the worst case scenario, to be restricted to two values.
After shifting and scaling these two values,  we can assume they are $\pm1$, without loss of generality.

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \cref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \protect\item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  \label{fig:models}}
\end{figure}

@@

The critical question is: what do these constraints imply about the space of achievable memory curves in \cref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \cref{eq:constr}?
We will derive some of these constraints in the next section.

@@

In the previous section, in \cref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[sec:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@
